Towards an Ecological Trait-data Standard

### 2017-12-12 manuscript/manuscript.Rmd
add comments of BKR and AO

Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets. 
To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on Biological Data; gfbio.org)  or database management systems for scientific projects (e.g. BExIS, XX ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.

After all, to mobilise the broad wealth of data into structured databases, it requires a certain minimum of information from the data providers. Getting accurate and well-defined input requires improving literacy of ecological researchers on data standards and ontologies.

@@

However, although open for manual download, the trait data on low-threshold repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and depending on commercial software. The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work. 
<!--To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or database management systems for scientific projects (e.g. BExIS, XX ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.--><!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information.

@@

Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories.

@@

<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. -->

@@

Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
More specifically, a trait dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an biological entity (i.e. an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.

@@

If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (Fig. 2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available. 
This format needs to clearly differentiate missing data and traits not applicable to a given taxon. It may store additional information (e.g. on variation of means or literature source) in secondary columns or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon).
Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14] and using Bayesian approaches [@schrodt15]. <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16].

@@

Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns. Each trait column represents one measured value. The width of the table varies with the number of traits included. Merging data from different sources requires manual harmonization of these structures.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Long-table dataset purport multiple advantages for the data manipulation (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets.

@@

We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management.

The online reference glossary of terms and description of the proposed structure is found at https://ecologicaltraitdata.github.io/ETS/.

The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

@@

As described above, tidy long-table formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories.

To enable a better standardisation of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names. In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. 
<!-- AG: This was a little bit confusing for me. In the manuscript, I understood that (for example) the taxon name element should be accompanied with a standard taxon name element which contains the URI to a web resource representing the (accepted) taxon. On the github page, both elements seem to be literals so that the “standard” element will just be used for a copy of the accepted name. In any case I would suggest to recommend to keep both the URI and the literal.-->
If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below). 
For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt').

@@

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09]. 
<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->

In some contexts, it may be safer to add such information directly as own columns of the data table to facilitate an analysis of co-factors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet. 
<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
The trade-off is data-consistency and readability *vs.* avoidance of content duplication.

Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample. Here, location or measurement information are in the primary focus of the investigation (see vocabulary extensions below). 
A general definition, whether a column describes asset data or is part of the central dataset is ill advised. Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context.

@@

It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits. This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively)<!--, according to a public ontology (e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass')-->.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11]. This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

@@

Many terms of these extensions refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).

@@

To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix. However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.

@@

Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added to the core data by calling the function `standardize.taxonomy()`. Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e. a thesaurus of traits. The same function harmonises trait values by converting values into target units and legit factor levels.

@@

We therefore encourage the development of trait thesauri in semantic ontologies. The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits. This effort is time consuming and lengthy, but it returns synergies and great scientific benefit by enabling a global synthesis, as the example of the global plant trait database TRY shows intricately [@diaz16].

>>>>

Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets. 
To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on Biological Data; gfbio.org)  or database <!-- AO: Just “data” or “research data”--> management systems for scientific projects (e.g. BEXIS, XX <!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.

After all, to mobilise the broad wealth of data into structured databases <!-- AO: Maybe better: formats-->, it requires a certain minimum of information from the original data providers. Getting accurate and well-defined input requires improving literacy of ecological researchers on data standards and ontologies.  <!-- AO: The last paragraphs may need more sources/proofings?-->

@@

However, although open for manual download, the trait data on low-threshold repositories might be stored in proprietary (e.g. .xlsx, .docx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious and depending on commercial software, putting the long-term accessibility of these data at risk. <!-- BKR: This might also result in data becoming no longer accessible (because the last computer on which the excel version required to read a file ran died -->
The column descriptions and applied terminologies for taxa and traits are rarely chosen to relate the data to other existing work. 
<!--To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or database management systems for scientific projects (e.g. BExIS, XX ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.--><!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->
With these constraints, low-threshold online repositories are a great source of open trait data, but combining such data requires much effort to standardize and harmonize information.

@@

Regarding open access, not all centralized databases for functional traits comply with the criteria demanded by journals and funding agencies, which leaves them at a disadvantage compared to low-threshold repositories. 
<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

@@

<!--Thus, in its simplest form,  a trait thesaurus would assign trait names with A) an unambiguous definition of the trait and B) an expected format (e.g. units or legit factor levels) of measured values or reported facts. It might additionally provide semantic relationships between terms for deriving a hierarchical or tree-based classification of traits. --><!-- BKR: Isn’t that the same? Or can you have multiple parents for one node? FDS: agree. simplify phrasing. -->

@@

Trait data have been defined  by @garnier17 to follow an entity-quality model (EQ), where a trait observation is phrased as 'an entity having a quality'. 
More specifically, a trait dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on a biological entity (i.e. an individual specimen, or parts of an individual) that can be assigned to a biological taxon (i.e. a species or higher-level taxon). 
Expaning on this definition, quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1*d*). 
The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a sub-sample or bodypart, an individual specimen, an entire species or higher-level taxon.

@@

<!-- AO: You argue somehow that matrix format I a result of lit. review and so on. I would not agree with it. Your example also “just” points to aggregated data which menas one row is on species (aggregated) value.
Your example is (for me) not really a matrix. In a matrix you have a relation between two variables. In the first column and and on the upper row normally the values of one variable are written. An example would be species x body length. Then for body_length some classifications are written like (0-1, 1-2, 2-3,…) Then in the matrix the observation/abundance/… is noted. E.g. spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on. Your wexample hat already different variables (=traits) with different units as columns. That is a meaningfull representation of variables and values in a normal data table. Maybe it is good her to show the enormous number of different investigated traits and point to the resulting big (many columns = wide) tables.
--><!--
If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait wide-table format (Fig. 2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available. 
This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g. '999'). It may store additional information (e.g. on variation of means or literature source) in secondary columns or accompanying datasheets. 
The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon). <!-- AO: You say it already 2 sentences before --><!--
Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14] and using Bayesian approaches [@schrodt15]. <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16]. --><!--AO: Why not? I can filter for a specific species and get all respective rows. But this is not a result a using a data format, it is a result of the data itself, FDS: no, since only one row per species exists; true that data itself don't cover this variation. -->
<!--

@@

Both the matrix format and the occurrence table format have the different trait values which have been collected or measured on one entity stored in multiple columns. Each trait column represents one value that applies to the entity in question (i.e. to a species in the first case or to an occurrence in the latter case). The width of the table varies with the number of traits included. Merging data from different sources requires user-defined mapping and manual harmonization of these structures.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr16], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). 
<!--AO: Useful for data sharing, ease of data harmonization, aggregation: probably yes. But not necessarily computational more effective at all. E.g. I start to mix up numeric values with descriptive values. I start to mix up different units. Now it is more complicated to retrieve associated values (like all length measurements in mm) again to make e.g. quality stuff. Ergo less effective from that perspective. But it has some advantages which may overrule dis. See below 
AO: For the whole proofing why to use such a long format I think that GBIF and DwC has already given some sentences. I strongly think their work should be considered and can be used as reference!
We can simply argue that a well established routine should now also executed on trait level!
Furthermore for DwC exists also the “extended measurement or facts extension” which is useful for putting e.g. traits to species.

--><!--
This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Long-table dataset purport multiple advantages for the data manipulation (e.g. filtering, sub-setting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modelling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets. 
<!-- BKR: Is that really true? Of course, you have less problems on the schema or structure level. But, on the other hand, the differences (e.g. different terms for the same trait,…) now happen on the instance level (that is – in the data; in the rows). Thus, at least with respect to integration, you just move the problem from one level to the other.  To ease integration, one would need to agree on the entries in the features column
AO: Maybe describe some more advantages here like: it is possible to ensure that the items/values are always taken from a provided lookup-table e.g. for taxon, trait_names, units, more?
We may have something like a dimension, the data source (measurement, lit. review, …) and some more criteria which are beneficial for researchers  and data aggregation?
Again: I think we find here papers regarding DWC and GBIF (and OBIS) as source of proofing the facts.
-->

@@

We were drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).  
We were particularly considering the work of @kattge11, and @garnier17, as well as @parr16 to ensure compatibility of datasets with major web-services and existing standards for biodiversity data management. <!-- AO: Webservices!! ;  Really to be considered as standards? -->

The online reference glossary of terms and description of the proposed structure can be found at https://ecologicaltraitdata.github.io/ETS/.

The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution<!-- BKR: rather it is meant as ... -->. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

@@

As described above, tidy long-table formats are the most predictable and compatible format for the core content of trait datasets and should be used for standalone uploads on low-threshold repositories. 
<!-- AO: Or rather respositories should provide a mechanism to “export” datasets into that proposed format
-->

To enable a better standardisation of nomenclatures and values, the entries of the core dataset should be related to semantic web references using Uniform Resource Identifiers (URIs) keep a duplicate record of accepted taxon names and standard trait names. In these duplicate columns, values should be reported as 'harmonized' entries, which are linked to ontologies and globally valid identifiers, consistent with semantic web standards [@berners-lee01; @wieczorek12; @page08]. <!-- BKR: There is some grammatical error here; not sure what you try to say--> 
<!-- AG: This was a little bit confusing for me. In the manuscript, I understood that (for example) the taxon name element should be accompanied with a standard taxon name element which contains the URI to a web resource representing the (accepted) taxon. On the github page, both elements seem to be literals so that the “standard” element will just be used for a copy of the accepted name. In any case I would suggest to recommend to keep both the URI and the literal.-->
If data contain multi-layered information on the measurement or occurrence (including geolocation and date information), further columns may be added to the core set (see description for Extensions below). 
For reasons of long-term accessibility, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt') with a unified character encoding (e.g. UTF-8). <!-- AO: Maybe less important  (everbody has access to an excel) but much more important is that it can be accessed also in future-->

@@

In very complex datasets repetition of verbose entries and long character strings in a single data table should be avoided [@wickham14]. Therefore we recommend use of  Darwin Core Archives [DwC-A, @robertson09]. 
<!-- AG: The refers to the “star schema” of DwC-Archive? You could perhaps mention that the redundancy can be reduced by semi-normalized formats such as x, y, z.-->
In some contexts, it may be safer to add such information directly as own columns of the data table to facilitate an analysis of co-factors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet. 
<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
The trade-off is data-consistency and readability *vs.* avoidance of content duplication.

Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample. Here, location or measurement information are in the primary focus of the investigation (see vocabulary extensions below). 
A general definition, whether a column describes asset data or is part of the central dataset is ill advised. Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context. 
<!-- FDS: shorten this and refer to existing standards for database use. -->

@@

It is indispensable to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits. This is achieved by adding globally valid identifiers for taxa (`taxonID`) and trait definitions (`traitID`), complemented by the more readable accepted names (`ScientificNameStd` and `traitNameStd`, respectively)<!--, according to a public ontology (e.g. GBIF Backbone Terminology; for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'; in TOP Thesaurus, the traitID for 'fruit mass' would be  'http://top-thesaurus.org/annotationInfo?viz=1&&trait=Fruit_mass')-->.  
By allowing for a double record of both user-specific and standardised entries, we acknowledge the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards  [as practised in TRY; @kattge11]. This redundancy of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. Keeping an additional reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms.  <!-- BKR: Why do you need this, if taxonIDs are used. They should be unambiguous
FDS: right, rephrase to: either use taxonID or provide this.-->
It is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15].

@@

Many terms of these extensions refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
<!-- BKR: Looking at your definitions, in most cases, I am not sure how you refine the terms. Is it really refinement or is it just using them with the same meaning they have in DwC?
-->

@@

To access data from public databases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases [including LEDA, BiolFlor and Ellenberg values; @bocci15] and aggregates them into a species-trait matrix. However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.

@@

Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added to the core data by calling the function `standardize.taxonomy()`. <!-- BKR: In the example below, this looks like a manual matching. What is done automatically here and which information does the user have to provide?  FDS: taxon is matched automatically, traits not. lack of trait APIs and webservices. --> 
Similarly, the `standardize.traits()` function matches user-provided trait names onto a given trait vocabulary and links it to an existing trait ontology or a user-provided table of trait definitions, i.e. a thesaurus of traits. The same function harmonises trait values by converting values into target units and legit factor levels. <!-- BKR: Have you thought about linking to unit ontologies here?
AO: This paper is a good reference and gives an idea of which ontology could be used;
http://www.semantic-web-journal.net/content/comparison-and-evaluation-ontologies-units-measurement
FDS: might be referred to in the section about thesauri -->

@@

We therefore encourage the development of trait thesauri in ontologies. The biggest challenge in community efforts of standardisation of traits may be the investment in consensus building which leads to an acceptance and establishment of the methodological and conceptual definitions of traits. This effort is time consuming and lengthy, but it returns synergies and great scientific benefit by enabling a global synthesis, as the example of the global plant trait database TRY shows intricately [@diaz16].
